Topological Equivariant Artist Model for Visualization Library Architecture

### 2025-01-06 paper/main.tex
clean up nc, pull back to equations, rework intro to equivariance

We construct the map as going from graphic to data because that encodes the notion that every visual element traces back to the data in some way, but there may be more data available than what is shown in the graphic. As exemplified in \autoref{fig:atct:morphisms:sheaf}, defining $\vindex$ as a surjective map allows allows us to express visual representations of a single record that are the union of many primitives, each of which has a different base space $\gbase_i$. These visual representations include multipart glyphs (e.g boxplots), combinations of plot types (e.g line with point markers), and the same point showing up in different continuities, such as $\dsection(\dbase_point)$ in \autoref{fig:atct:morphisms:sheaf} and a station in \autoref{fig:related-work:continuity:ktypes}.

@@

The commutativity of \autoref{eq:artist:homset} means that an artist over data space $\vartist_{\dbase}: \dsection \mapsto \vindexpush \gsection$, an artist over graphic space $vartist_{\gbase}: \vindexpull \dsection \mapsto \gsection$, and an artist $\vartist: \dsection \mapsto \gsection$ are equivalent such that:
\begin{equation*}
  \begin{split}
  \dsection(\dbasepoint) &= \vindexpull\dsection(\gbasepoint)  \\
   \implies
  \vartist_{\dbase}(\dsection(\dbasepoint)) &= \vartist_{\gbase}(\vindexpull\dsection(\gbasepoint)) = \vartist(\dsection(\dbasepoint))\\
  \implies \vindexpush\gsection(\gbasepoint) &= \gsection(\gbasepoint)
  \end{split}
\end{equation*}
when $\vindex(\gbasepoint) = \dbasepoint$. The ability to shift base spaces using adaptors (\vindexpush, \vindexpull) allows a developer to specify structure preserving visualization components in data space, graphic space, or a mix of both depending on what is optimal for the specific use case. For example, mapping a line to a color in data space, but drawing the line in graphic space so that it can be dynamically resampled.

Because Artists can be constructed as morphisms of sheaves over the same base spaces through the application of pushforward and pullback functors, they are natural transformations \cite{bradleyWhatNaturalTransformation}.
\begin{definition}\label{def:natural-transform}
  Given two functors $F, G: \mathcal{C}\rightarrow \mathcal{D}$, a \textbf{natural transformation} $\alpha: F \rightarrow G$ is a function which assigns to each object $c$ of $\mathcal{C}$ a morphism $\alpha_c:F(input) \rightarrow G(c), G(c) \in \mathcal{D}$, in such a way that for every morphism $f:c \rightarrow c^\prime, c^\prime \in \mathcal{C}$, the morphisms in $\mathcal{D}$ commute such that $\alpha_c^{\prime}(F(f)(F(c))) = G(f)(\alpha_c(F(c))$. When this holds, $\alpha_{c}$ is \textit{natural} in $c$.\cite{maclaneCategoriesWorkingMathematician2013}.
\end{definition}.

This means that the structure on $\mathcal{C}$, as defined by its input objects $c \in \mathcal{C}$ and morphisms $f: c \rightarrow c^\prime$ are preserved in the output of the functors $F, G$ and by maps between functors $\alpha$. This preservation of the structure of the input in the output encapsulates the structure preservation discussed in \autoref{sec:related-work:equivariance} and serves as the basis for describing structure preservation in \autoref{sec:artist:equiv}. Because the Artist is a natural transform and natural transforms are maps of functors that take the same input object and return objects in the same category\cite{milewskiCategoryTheoryProgrammers}, the artist signature can be formulated as:
 \begin{align*}
  \textcolor{artist}{\texttt{Artist}}:&(\textcolor{section}{\texttt{dataset}}: \textcolor{base}{\texttt{topology}}\textcolor{section}{\rightarrow} \textcolor{fiber}{\texttt{fields}}) \\
  \textcolor{artist}{\rightarrow} & (\textcolor{section}{\texttt{graphic}}: \textcolor{base}{\texttt{topology}}\textcolor{section}{\rightarrow} \textcolor{fiber}{\texttt{fields}})
 \end{align*}
where \texttt{topology} and \texttt{field} are meta classes of compatible types.

@@

The artist is specified as a natural natural transform (\autoref{def:natural-transform}) on sheaves because it means that the artist can assume that subsets of the data behave in the same manner as the whole when the artist preserves the presheaf and sheaf constraints (\autoref{def:atct:presheaf}, \autoref{def:atct:sheaf}) when mapping data into graphics. Preserving the presheaf constraints means that the set of possible visualizations for a larger set of values must include all visualizations for a subset of those values  $\vartist(\Gamma(\dbase_1, \dtotal)) \xrightarrow{i^*} \vartist(\Gamma(\dbase_2, \dtotal))$ when more values may exist on the subset of the index $\dbase_2 \xhookrightarrow{i} \dbase_1$. While the curves in \autoref{fig:atct:presheaf} are meant as stand-ins for the data, they also serve to illustrate that the relationship is identical in the graphic space. The artist preserves the \textit{locality} axiom in that identical datasets $\dsection^{a} = \dsection^{b}$ generate identical specifications $\vartist(\dsection^a) = \vartist(\dsection^b)$. The artist respects \textit{gluing} in that graphics are composable, as described in \autoref{sec:artist:operators}, such that a set of artists that generate graphics for parts of a dataset are equivalent to an artist that generates a graphic for the whole dataset. Managing subsets of data and on demand data in a structure preserving way is necessary to support visualizing streaming data \cite{krstajicVisualizationStreamingData2013} and techniques that rely on shifting or resampling the indexing space, such as  panning and zooming \cite{NekrasovskiEvaluationPanZoom2006}.

\subsection{Equivariance}
\label{sec:artist:equiv}
As introduced in \autoref{sec:related-work:equivariance}, data and the corresponding visual encoding are expected to have compatible structure. This structure can be formally expressed as actions $\dfunc \in \dfuncset$ on the sheaf $\sheaf_{\dbase, \dtotal}$. We generalize from binary operations to a family of actions because that allows for expanding the set of allowable transformations on the data beyond a single operator. We describe the changes on the graphic side as changes in measurements \measure\, which are scaler or vector components of the rendered graphic that can be quantified, such as the color, position, shape, texture, or rotation angle of the graphic. The visual variables \cite{bertinIIPropertiesGraphic2011} are a subset of measurable components. For example, a measurement of a scatter marker could be its color (e.g. red) or its x position (e.g. 5).

>>>>

We construct the map as going from graphic to data because that encodes the notion that every visual element traces back to the data in some way, but there may be more data available than what is shown in the graphic. As exemplified in \autoref{fig:atct:morphisms:sheaf}, defining $\vindex$ as a surjective map allows allows us to express visual representations of a single record that are the union of many primitives, each of which has a different base space $\gbase_i$. These visual representations include multipart glyphs (e.g boxplots), combinations of plot types (e.g line with point markers), and the same point showing up in different continuities, such as $\dsection(\dbase_{point})$ in \autoref{fig:atct:morphisms:sheaf} and a station in \autoref{fig:related-work:continuity:ktypes}.

@@

The commutativity of \autoref{eq:artist:homset} means that an artist over data space $\vartist_{\dbase}: \dsection \mapsto \vindexpush \gsection$, an artist over graphic space $\vartist_{\gbase}: \vindexpull \dsection \mapsto \gsection$, and an artist $\vartist: \dsection \mapsto \gsection$ are equivalent. Specifically, given \autoref{eq:atct:sheaf:pullback_hover} (\vindexpull) and \autoref{eq:atct:sheaf:pushforward_select} (\vindexpush), $\vartist_{\dbase}(\dsection(\dbasepoint)) = \vartist_{\gbase}(\vindexpull\dsection(\gbasepoint)) = \vartist(\dsection(\dbasepoint)$ when $\vindex(\gbasepoint) = \dbasepoint$. The ability to shift base spaces using adaptors (\vindexpush, \vindexpull) allows a developer to specify structure preserving visualization components in data space, graphic space, or a mix of both depending on what is optimal for the specific use case. For example, mapping a line to a color in data space, but drawing the line in graphic space so that it can be dynamically resampled based on panning and zooming.

Because Artists can be constructed as morphisms of sheaves over either \dbase\ or \gbase\ through the application of pushforward and pullback functors (\vindexpush, \vindexpull), they are natural transformations \cite{bradleyWhatNaturalTransformation}.
\begin{definition}\label{def:natural-transform}
  Given two functors $F, G: \mathcal{C}\rightarrow \mathcal{D}$, a \textbf{natural transformation} $\alpha: F \rightarrow G$ is a function which assigns to each object $c$ of $\mathcal{C}$ a morphism $\alpha_c:F(input) \rightarrow G(c), G(c) \in \mathcal{D}$, in such a way that for every morphism $f:c \rightarrow c^\prime, c^\prime \in \mathcal{C}$, the morphisms in $\mathcal{D}$ commute such that $\alpha_c^{\prime}(F(f)(F(c))) = G(f)(\alpha_c(F(c))$. When this holds, $\alpha_{c}$ is \textit{natural} in $c$.\cite{maclaneCategoriesWorkingMathematician2013}.
\end{definition}.

This means that the structure on $\mathcal{C}$, as defined by its input objects $c \in \mathcal{C}$ and morphisms $f: c \rightarrow c^\prime$ are preserved in the output of the functors $F, G$ and by maps between functors $\alpha$. The artist \vartist\ is the natural transform mapping the sheaf functors $\sheaf_{data} \rightarrow \sheaf_{graphic}$, which both have the signature $\dbase^{op} (\gbase^{op})\rightarrow \texttt{set}$ \autoref{def:atct:presheaf}. This preservation of the structure of the input in the output encapsulates the structure preservation discussed in \autoref{sec:related-work:equivariance} and serves as the basis for describing structure preservation in \autoref{sec:artist:equiv}. Because the Artist is a natural transform and natural transforms are maps of functors that take the same input object and return objects in the same category\cite{milewskiCategoryTheoryProgrammers}, the artist signature can be formulated as:
 \begin{align*}
  \textcolor{artist}{\texttt{Artist}}:&(\textcolor{section}{\texttt{dataset}}: \textcolor{base}{\texttt{topology}}\textcolor{section}{\rightarrow} \textcolor{fiber}{\texttt{fields}}) \\
  \textcolor{artist}{\rightarrow} & (\textcolor{section}{\texttt{graphic}}: \textcolor{base}{\texttt{topology}}\textcolor{section}{\rightarrow} \textcolor{fiber}{\texttt{fields}})
 \end{align*}
where \texttt{topology} and \texttt{field} are meta classes of compatible types.

@@

Because the artist is specified as a natural transform (\autoref{def:natural-transform}) on sheaves, the artist expects that subsets of the dataset have the same properties as the whole dataset when the artist preserves the presheaf and sheaf constraints (\autoref{def:atct:presheaf}, \autoref{def:atct:sheaf}) when mapping data into graphics. Preserving the presheaf constraints means that the set of possible visualizations for a larger set of values must include all visualizations for a subset of those values  $\vartist(\Gamma(\dbase_1, \dtotal)) \xrightarrow{i^*} \vartist(\Gamma(\dbase_2, \dtotal))$ when more values may exist on the subset of the index $\dbase_2 \xhookrightarrow{i} \dbase_1$. While the curves in \autoref{fig:atct:presheaf} are meant as stand-ins for the data, they also serve to illustrate that the relationship is identical in the graphic space. The artist preserves the \textit{locality} axiom in that identical datasets $\dsection^{a} = \dsection^{b}$ generate identical specifications $\vartist(\dsection^a) = \vartist(\dsection^b)$. The artist respects \textit{gluing} in that graphics are composable, as described in \autoref{sec:artist:operators}, such that a set of artists that generate graphics for parts of a dataset are equivalent to an artist that generates a graphic for the whole dataset. Managing subsets of data and on demand data in a structure preserving way is necessary to support visualizing streaming data \cite{krstajicVisualizationStreamingData2013} and techniques that rely on shifting or resampling the indexing space, such as  panning and zooming \cite{NekrasovskiEvaluationPanZoom2006}.

\subsection{Equivariance}
\label{sec:artist:equiv}
As introduced in \autoref{sec:related-work:equivariance}, data and the corresponding visual encoding are expected to have compatible structure such that actions (\autoref{def:related-work:action}) on data and graphic are symmetric. We generalize from binary operations to a family of actions $\dfunc \in \dfuncset$ on the sheaf $\sheaf_{\dbase, \dtotal}$ because that allows for expanding the set of allowable transformations on the data beyond a single operator. We describe the changes on the graphic side as changes in measurements \measure\ of visual components. A superset of the Bertin visual variables \cite{\cite{bertinIIPropertiesGraphic2011}, measurable components are scaler or vector components of the rendered graphic that can be quantified, such as the color, position, shape, texture, or rotation angle of the graphic. For example, a measurement of a scatter marker could be its color (e.g. red) or its x position (e.g. 5).
